quollr: An R Package for Visualizing 2-D Models from Non-linear Dimension Reductions in High Dimensional Space

### 2025-09-18 paper-quollr.Rmd
Updated notations

This function requires several parameters: the high-dimensional data (`highd_data`), the emdedding data (`nldr_data`), the number of bins along the x-axis (`b1`), the buffer amount as a proportion of data (`q`), and benchmark value to extract high density hexagons (`hd_thresh`). The function returns an object (`nldr_scaled_obj`) containing three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); the hexagonal object (`hb_obj`), the fitted model in both $2\text{-}D$ (`model_2d`), and $p\text{-}D$ (`model_highd`), and triangular mesh (`trimesh_data`).

@@

Each hexagon is defined relative to its centroid $(C_x, C_y)$, with six vertices positioned equidistantly around the center. The function first verifies the presence of the required hexagon width parameter `a1`. This width determines the horizontal spacing.

Two derived constants are calculated to define the relative distances to the vertices. The horizontal and vertical offset is defined as $dx = a_1/2$, and $dy = a_1/\sqrt{3}$ repectively. A vertical spacing factor $v = a_1/2\sqrt{3}$ refines vertical placement in staggered rows.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The `assign_data()` function performs this assignment by calculating the $2\text{-}D$ Euclidean distance between each point in the $2\text{-}D$ embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the $2\text{-}D$ layout. It then selects the corresponding x and y coordinates of each hexagon’s centroid.

@@

The result is a data frame of the scaled $2\text{-}D$ embedding with an additional `h` column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by `h`, which represents the hexagon label associated with each point in the $2\text{-}D$ layout. Within each group, it collects the `ID`s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, `pts_list`, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a $2\text{-}D$ layout, we employ Delaunay triangulation on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or "holes" in the $2\text{-}D$ structure, potentially disrupting the continuity of the representation.

@@

The final step involves lifting the fitted \twoD model into \pD. This is done by modelling a point in \pD as the  \pD mean of data points in the \twoD centroid. This is performed using the `avg_highd_data()` function, which takes $p\text{-}D$ data (`highd_data`) and embedding data with their corresponding hexagonal bin IDs as inputs (`scaled_nldr_hexid`).

@@

The `predict_emb()` function is used to predict a point in a $2\text{-}D$ embedding for a new $p\text{-}D$ data point using the fitted model. This function is useful to predict $2\text{-}D$ embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest $p\text{-}D$ model point is identified for the new $p\text{-}D$ data point by computing $p\text{-}D$ Euclidean distance. Then, the corresponding $2\text{-}D$ bin centroid mapping for the identified $p\text{-}D$ model point is determined. Finally, the coordinates of the identified $2\text{-}D$ bin centroid is used as the predicted NLDR embedding for the new $p\text{-}D$ data point.

@@

Furthermore, `augment()` accepts $2\text{-}D$ and $p\text{-}D$ model points, and the $p\text{-}D$ data and adds information about each observation in the data set.

@@

The package offers several $2\text{-}D$ visualizations, including:

@@

The generated $p\text{-}D$ model, overlaid with the data, can also be visualized using `show_langevitour`. Additionally, it features a function for visualizing the $2\text{-}D$ projection of the fitted model overlaid on the data, called `plot_proj`.

@@

The `geom_trimesh()` function introduces a custom `ggplot2` layer designed for visualizing $2\text{-}D$ wireframe on a provided set of bin centroids.

@@

#### $p\text{-}D$ model visualization

To visualise how well the $p\text{-}D$ model captures the underlying structure of the high-dimensional data, we provide a tour of the model in \pD using the `show_langevitour()` function. This function renders a dynamic projection of both the high-dimensional data and the model using the `langevitour` package.

@@

In the resulting interactive visualization black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the $2\text{-}D$ wireframe structure mapped back to high-dimensional space.

@@

To support interactive evaluation of how well the $p\text{-}D$ model captures the structure of the high-dimensional data, we introduce `show_link_plots()`. This visualization combines two complementary views: the nonlinear dimension reduction (NLDR) representation and a dynamic tour of the model ovelaid the data in the high-dimensional space. Both views are interactively linked, enabling users to explore.

Before visualization, the input data must be prepared using the `comb_all_data_model()` function. This function combines the high-dimensional data (`highd_data`), the NLDR data (`nldr_data`), and the bin-averaged high-dimensional model representation (`model_highd`) aligned to the $2\text{-}D$ bin layout (`model_2d`):

@@

The function `show_link_plots()` generates two side-by-side, interactively linked plots; a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Before visualization, the input data must be prepared using the `comb_all_data_model_error()` function. The function requires several arguments: points data which contain high-dimensional data (`highd_data`), NLDR data (`nldr_data`), high-dimensional model data (`model_highd`), $2\text{-}D$ model data (`model_2d`), and model error (`error_data`).

@@

The function `show_error_link_plots()` generates three side-by-side, interactively linked plots; a error distribution, a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model_error()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Several core computations within `quollr` are optimized using compiled C++ code via the `Rcpp` and `RcppArmadillo` packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (`compute_highd_dist()`), error metrics (`compute_errors()`), $2\text{-}D$ distance calculations (`calc_2d_dist_cpp()`), and generation of hexagon coordinates (`gen_hex_coord_cpp()`) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \@ref(fig:limb-rmse) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better $2\text{-}D$ layout.

@@

The `quollr` package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in $2\text{-}D$ and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a $2\text{-}D$ layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate $2\text{-}D$ representations.

@@

To support efficient computation, particularly for large-scale datasets, several core functions in `quollr` are implemented in C++ using `Rcpp` and `RcppArmadillo`. These include functions for computing Euclidean distances in high-dimensional and $2\text{-}D$ space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, `compute_highd_dist()` accelerates nearest neighbor lookup in high-dimensional space, `compute_errors()` calculates RMSE and total absolute error efficiently, and `calc_2d_dist_cpp()` speeds up distance calculations in $2\text{-}D$. Additionally, `gen_hex_coord_cpp()` constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.

>>>>

This function requires several parameters: the high-dimensional data (`highd_data`), the emdedding data (`nldr_data`), the number of bins along the x-axis (`b1`), the buffer amount as a proportion of data (`q`), and benchmark value to extract high density hexagons (`hd_thresh`). The function returns an object (`nldr_scaled_obj`) containing three elements: the first is the scaled NLDR data (`scaled_nldr`), and the second and third are the limits of the original NLDR data (`lim1` and `lim2`); the hexagonal object (`hb_obj`), the fitted model in both \twoD (`model_2d`), and \pD (`model_highd`), and triangular mesh (`trimesh_data`).

@@

Each hexagon is defined relative to its centroid $(C_x, C_y)$, with six vertices positioned equidistantly around the center. The function first verifies the presence of the required hexagon width parameter `a1`. This width determines the horizontal spacing (`hs`).

Two derived constants are calculated to define the relative distances to the vertices. The horizontal and vertical offset is defined as $dx = a_1/2$, and $dy = a_1/\sqrt{3}$ repectively. A vertical spacing factor $vf = a_1/2\sqrt{3}$ refines vertical placement in staggered rows.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The `assign_data()` function performs this assignment by calculating the \twoD Euclidean distance between each point in the \twoD embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the \twoD layout. It then selects the corresponding x and y coordinates of each hexagon’s centroid.

@@

The result is a data frame of the scaled \twoD embedding with an additional `h` column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by `h`, which represents the hexagon label associated with each point in the \twoD layout. Within each group, it collects the `ID`s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, `pts_list`, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a \twoD layout, we employ Delaunay triangulation on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or "holes" in the \twoD structure, potentially disrupting the continuity of the representation.

@@

The final step involves lifting the fitted \twoD model into \pD. This is done by modelling a point in \pD as the  \pD mean of data points in the \twoD centroid. This is performed using the `avg_highd_data()` function, which takes \pD data (`highd_data`) and embedding data with their corresponding hexagonal bin IDs as inputs (`scaled_nldr_hexid`).

@@

The `predict_emb()` function is used to predict a point in a \twoD embedding for a new \pD data point using the fitted model. This function is useful to predict \twoD embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest \pD model point is identified for the new \pD data point by computing \pD Euclidean distance. Then, the corresponding \twoD bin centroid mapping for the identified \pD model point is determined. Finally, the coordinates of the identified \twoD bin centroid is used as the predicted NLDR embedding for the new \pD data point.

@@

Furthermore, `augment()` accepts \twoD and \pD model points, and the \pD data and adds information about each observation in the data set.

@@

The package offers several \twoD visualizations, including:

@@

The generated \pD model, overlaid with the data, can also be visualized using `show_langevitour`. Additionally, it features a function for visualizing the \twoD projection of the fitted model overlaid on the data, called `plot_proj`.

@@

The `geom_trimesh()` function introduces a custom `ggplot2` layer designed for visualizing \twoD wireframe on a provided set of bin centroids.

@@

#### \pD model visualization

To visualise how well the \pD model captures the underlying structure of the high-dimensional data, we provide a tour of the model in \pD using the `show_langevitour()` function. This function renders a dynamic projection of both the high-dimensional data and the model using the `langevitour` package.

@@

In the resulting interactive visualization black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the \twoD wireframe structure mapped back to high-dimensional space.

@@

To support interactive evaluation of how well the \pD model captures the structure of the high-dimensional data, we introduce `show_link_plots()`. This visualization combines two complementary views: the nonlinear dimension reduction (NLDR) representation and a dynamic tour of the model ovelaid the data in the high-dimensional space. Both views are interactively linked, enabling users to explore.

Before visualization, the input data must be prepared using the `comb_all_data_model()` function. This function combines the high-dimensional data (`highd_data`), the NLDR data (`nldr_data`), and the bin-averaged high-dimensional model representation (`model_highd`) aligned to the \twoD bin layout (`model_2d`):

@@

The function `show_link_plots()` generates two side-by-side, interactively linked plots; a \twoD NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Before visualization, the input data must be prepared using the `comb_all_data_model_error()` function. The function requires several arguments: points data which contain high-dimensional data (`highd_data`), NLDR data (`nldr_data`), high-dimensional model data (`model_highd`), \twoD model data (`model_2d`), and model error (`error_data`).

@@

The function `show_error_link_plots()` generates three side-by-side, interactively linked plots; a error distribution, a \twoD NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the `langevitour` package), displaying both the data and the model. The function takes the output from `comb_all_data_model_error()` (`point_data`) and `edge_data` which defines connections between neighboring bins.

@@

Several core computations within `quollr` are optimized using compiled C++ code via the `Rcpp` and `RcppArmadillo` packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (`compute_highd_dist()`), error metrics (`compute_errors()`), \twoD distance calculations (`calc_2d_dist_cpp()`), and generation of hexagon coordinates (`gen_hex_coord_cpp()`) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

Limb muscle cells of mice in @tabula2018 are examined. There are $1067$ single cells, with $14997$ gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \@ref(fig:limb-rmse) (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better \twoD layout.

@@

The `quollr` package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in \twoD and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a \twoD layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate \twoD representations.

@@

To support efficient computation, particularly for large-scale datasets, several core functions in `quollr` are implemented in C++ using `Rcpp` and `RcppArmadillo`. These include functions for computing Euclidean distances in high-dimensional and \twoD space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, `compute_highd_dist()` accelerates nearest neighbor lookup in high-dimensional space, `compute_errors()` calculates RMSE and total absolute error efficiently, and `calc_2d_dist_cpp()` speeds up distance calculations in \twoD. Additionally, `gen_hex_coord_cpp()` constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.

### 2025-09-18 paper-quollr.tex
Updated notations

This function requires several parameters: the high-dimensional data (\texttt{highd\_data}), the emdedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{b1}), the buffer amount as a proportion of data (\texttt{q}), and benchmark value to extract high density hexagons (\texttt{hd\_thresh}). The function returns an object (\texttt{nldr\_scaled\_obj}) containing three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); the hexagonal object (\texttt{hb\_obj}), the fitted model in both \(2\text{-}D\) (\texttt{model\_2d}), and \(p\text{-}D\) (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

@@

Each hexagon is defined relative to its centroid \((C_x, C_y)\), with six vertices positioned equidistantly around the center. The function first verifies the presence of the required hexagon width parameter \texttt{a1}. This width determines the horizontal spacing.

Two derived constants are calculated to define the relative distances to the vertices. The horizontal and vertical offset is defined as \(dx = a_1/2\), and \(dy = a_1/\sqrt{3}\) repectively. A vertical spacing factor \(v = a_1/2\sqrt{3}\) refines vertical placement in staggered rows.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The \texttt{assign\_data()} function performs this assignment by calculating the \(2\text{-}D\) Euclidean distance between each point in the \(2\text{-}D\) embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the \(2\text{-}D\) layout. It then selects the corresponding x and y coordinates of each hexagon's centroid.

@@

The result is a data frame of the scaled \(2\text{-}D\) embedding with an additional \texttt{h} column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by \texttt{h}, which represents the hexagon label associated with each point in the \(2\text{-}D\) layout. Within each group, it collects the \texttt{ID}s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, \texttt{pts\_list}, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a \(2\text{-}D\) layout, we employ Delaunay triangulation on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or ``holes'' in the \(2\text{-}D\) structure, potentially disrupting the continuity of the representation.

@@

The final step involves lifting the fitted $2\text{-}D$ model into $p\text{-}D$. This is done by modelling a point in $p\text{-}D$ as the $p\text{-}D$ mean of data points in the $2\text{-}D$ centroid. This is performed using the \texttt{avg\_highd\_data()} function, which takes \(p\text{-}D\) data (\texttt{highd\_data}) and embedding data with their corresponding hexagonal bin IDs as inputs (\texttt{scaled\_nldr\_hexid}).

@@

The \texttt{predict\_emb()} function is used to predict a point in a \(2\text{-}D\) embedding for a new \(p\text{-}D\) data point using the fitted model. This function is useful to predict \(2\text{-}D\) embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest \(p\text{-}D\) model point is identified for the new \(p\text{-}D\) data point by computing \(p\text{-}D\) Euclidean distance. Then, the corresponding \(2\text{-}D\) bin centroid mapping for the identified \(p\text{-}D\) model point is determined. Finally, the coordinates of the identified \(2\text{-}D\) bin centroid is used as the predicted NLDR embedding for the new \(p\text{-}D\) data point.

@@

Furthermore, \texttt{augment()} accepts \(2\text{-}D\) and \(p\text{-}D\) model points, and the \(p\text{-}D\) data and adds information about each observation in the data set.

@@

The package offers several \(2\text{-}D\) visualizations, including:

@@

The generated \(p\text{-}D\) model, overlaid with the data, can also be visualized using \texttt{show\_langevitour}. Additionally, it features a function for visualizing the \(2\text{-}D\) projection of the fitted model overlaid on the data, called \texttt{plot\_proj}.

@@

The \texttt{geom\_trimesh()} function introduces a custom \texttt{ggplot2} layer designed for visualizing \(2\text{-}D\) wireframe on a provided set of bin centroids.

@@

\paragraph{\texorpdfstring{\(p\text{-}D\) model visualization}{p\textbackslash text\{-\}D model visualization}}\label{ptext-d-model-visualization}

To visualise how well the \(p\text{-}D\) model captures the underlying structure of the high-dimensional data, we provide a tour of the model in $p\text{-}D$ using the \texttt{show\_langevitour()} function. This function renders a dynamic projection of both the high-dimensional data and the model using the \texttt{langevitour} package.

@@

In the resulting interactive visualization black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the \(2\text{-}D\) wireframe structure mapped back to high-dimensional space.

@@

To support interactive evaluation of how well the \(p\text{-}D\) model captures the structure of the high-dimensional data, we introduce \texttt{show\_link\_plots()}. This visualization combines two complementary views: the nonlinear dimension reduction (NLDR) representation and a dynamic tour of the model ovelaid the data in the high-dimensional space. Both views are interactively linked, enabling users to explore.

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model()} function. This function combines the high-dimensional data (\texttt{highd\_data}), the NLDR data (\texttt{nldr\_data}), and the bin-averaged high-dimensional model representation (\texttt{model\_highd}) aligned to the \(2\text{-}D\) bin layout (\texttt{model\_2d}):

@@

The function \texttt{show\_link\_plots()} generates two side-by-side, interactively linked plots; a \(2\text{-}D\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model\_error()} function. The function requires several arguments: points data which contain high-dimensional data (\texttt{highd\_data}), NLDR data (\texttt{nldr\_data}), high-dimensional model data (\texttt{model\_highd}), \(2\text{-}D\) model data (\texttt{model\_2d}), and model error (\texttt{error\_data}).

@@

The function \texttt{show\_error\_link\_plots()} generates three side-by-side, interactively linked plots; a error distribution, a \(2\text{-}D\) NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model\_error()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Several core computations within \texttt{quollr} are optimized using compiled C++ code via the \texttt{Rcpp} and \texttt{RcppArmadillo} packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (\texttt{compute\_highd\_dist()}), error metrics (\texttt{compute\_errors()}), \(2\text{-}D\) distance calculations (\texttt{calc\_2d\_dist\_cpp()}), and generation of hexagon coordinates (\texttt{gen\_hex\_coord\_cpp()}) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \ref{fig:limb-rmse} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better \(2\text{-}D\) layout.

@@

The \texttt{quollr} package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in \(2\text{-}D\) and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a \(2\text{-}D\) layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate \(2\text{-}D\) representations.

@@

To support efficient computation, particularly for large-scale datasets, several core functions in \texttt{quollr} are implemented in C++ using \texttt{Rcpp} and \texttt{RcppArmadillo}. These include functions for computing Euclidean distances in high-dimensional and \(2\text{-}D\) space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, \texttt{compute\_highd\_dist()} accelerates nearest neighbor lookup in high-dimensional space, \texttt{compute\_errors()} calculates RMSE and total absolute error efficiently, and \texttt{calc\_2d\_dist\_cpp()} speeds up distance calculations in \(2\text{-}D\). Additionally, \texttt{gen\_hex\_coord\_cpp()} constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.

>>>>

This function requires several parameters: the high-dimensional data (\texttt{highd\_data}), the emdedding data (\texttt{nldr\_data}), the number of bins along the x-axis (\texttt{b1}), the buffer amount as a proportion of data (\texttt{q}), and benchmark value to extract high density hexagons (\texttt{hd\_thresh}). The function returns an object (\texttt{nldr\_scaled\_obj}) containing three elements: the first is the scaled NLDR data (\texttt{scaled\_nldr}), and the second and third are the limits of the original NLDR data (\texttt{lim1} and \texttt{lim2}); the hexagonal object (\texttt{hb\_obj}), the fitted model in both $2\text{-}D$ (\texttt{model\_2d}), and $p\text{-}D$ (\texttt{model\_highd}), and triangular mesh (\texttt{trimesh\_data}).

@@

Each hexagon is defined relative to its centroid \((C_x, C_y)\), with six vertices positioned equidistantly around the center. The function first verifies the presence of the required hexagon width parameter \texttt{a1}. This width determines the horizontal spacing (\texttt{hs}).

Two derived constants are calculated to define the relative distances to the vertices. The horizontal and vertical offset is defined as \(dx = a_1/2\), and \(dy = a_1/\sqrt{3}\) repectively. A vertical spacing factor \(vf = a_1/2\sqrt{3}\) refines vertical placement in staggered rows.

@@

After generating the centroids that define the hexagonal grid, the next step is to assign each point in the NLDR embedding to its nearest hexagonal bin. The \texttt{assign\_data()} function performs this assignment by calculating the $2\text{-}D$ Euclidean distance between each point in the $2\text{-}D$ embedding and all hexagon centroids.

First, the function extracts the first two dimensions of the scaled NLDR embedding, which represent the $2\text{-}D$ layout. It then selects the corresponding x and y coordinates of each hexagon's centroid.

@@

The result is a data frame of the scaled $2\text{-}D$ embedding with an additional \texttt{h} column, indicating the hexagonal bin to which each point belongs.

@@

The function first groups the input data by \texttt{h}, which represents the hexagon label associated with each point in the $2\text{-}D$ layout. Within each group, it collects the \texttt{ID}s into a list, resulting in a summary data frame where each row corresponds to a single hexagon. The resulting column, \texttt{pts\_list}, contains all point identifiers associated with that hexagon.

@@

To represent the neighborhood structure of hexagonal bins in a $2\text{-}D$ layout, we employ Delaunay triangulation on the centroids of hexagons. This geometric approach is used to infer which bins are considered neighbors.

@@

Not all hexagons contain meaningful information. Some may have very few or no data points due to the sparsity or shape of the underlying structure. Simply removing hexagons with low counts (e.g., fewer than a fixed threshold) can lead to gaps or ``holes'' in the $2\text{-}D$ structure, potentially disrupting the continuity of the representation.

@@

The final step involves lifting the fitted $2\text{-}D$ model into $p\text{-}D$. This is done by modelling a point in $p\text{-}D$ as the $p\text{-}D$ mean of data points in the $2\text{-}D$ centroid. This is performed using the \texttt{avg\_highd\_data()} function, which takes $p\text{-}D$ data (\texttt{highd\_data}) and embedding data with their corresponding hexagonal bin IDs as inputs (\texttt{scaled\_nldr\_hexid}).

@@

The \texttt{predict\_emb()} function is used to predict a point in a $2\text{-}D$ embedding for a new $p\text{-}D$ data point using the fitted model. This function is useful to predict $2\text{-}D$ embedding irrespective of the NLDR technique.

In the prediction process, first, the nearest $p\text{-}D$ model point is identified for the new $p\text{-}D$ data point by computing $p\text{-}D$ Euclidean distance. Then, the corresponding $2\text{-}D$ bin centroid mapping for the identified $p\text{-}D$ model point is determined. Finally, the coordinates of the identified $2\text{-}D$ bin centroid is used as the predicted NLDR embedding for the new $p\text{-}D$ data point.

@@

Furthermore, \texttt{augment()} accepts $2\text{-}D$ and $p\text{-}D$ model points, and the $p\text{-}D$ data and adds information about each observation in the data set.

@@

The package offers several $2\text{-}D$ visualizations, including:

@@

The generated $p\text{-}D$ model, overlaid with the data, can also be visualized using \texttt{show\_langevitour}. Additionally, it features a function for visualizing the $2\text{-}D$ projection of the fitted model overlaid on the data, called \texttt{plot\_proj}.

@@

The \texttt{geom\_trimesh()} function introduces a custom \texttt{ggplot2} layer designed for visualizing $2\text{-}D$ wireframe on a provided set of bin centroids.

@@

\paragraph{\texorpdfstring{$p\text{-}D$ model visualization}{model visualization}}\label{model-visualization}

To visualise how well the $p\text{-}D$ model captures the underlying structure of the high-dimensional data, we provide a tour of the model in $p\text{-}D$ using the \texttt{show\_langevitour()} function. This function renders a dynamic projection of both the high-dimensional data and the model using the \texttt{langevitour} package.

@@

In the resulting interactive visualization black points represent the high-dimensional data, green points represent the model centroids from each bin, and the lines between model points reflect the $2\text{-}D$ wireframe structure mapped back to high-dimensional space.

@@

To support interactive evaluation of how well the $p\text{-}D$ model captures the structure of the high-dimensional data, we introduce \texttt{show\_link\_plots()}. This visualization combines two complementary views: the nonlinear dimension reduction (NLDR) representation and a dynamic tour of the model ovelaid the data in the high-dimensional space. Both views are interactively linked, enabling users to explore.

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model()} function. This function combines the high-dimensional data (\texttt{highd\_data}), the NLDR data (\texttt{nldr\_data}), and the bin-averaged high-dimensional model representation (\texttt{model\_highd}) aligned to the $2\text{-}D$ bin layout (\texttt{model\_2d}):

@@

The function \texttt{show\_link\_plots()} generates two side-by-side, interactively linked plots; a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Before visualization, the input data must be prepared using the \texttt{comb\_all\_data\_model\_error()} function. The function requires several arguments: points data which contain high-dimensional data (\texttt{highd\_data}), NLDR data (\texttt{nldr\_data}), high-dimensional model data (\texttt{model\_highd}), $2\text{-}D$ model data (\texttt{model\_2d}), and model error (\texttt{error\_data}).

@@

The function \texttt{show\_error\_link\_plots()} generates three side-by-side, interactively linked plots; a error distribution, a $2\text{-}D$ NLDR representation, and a dynamic projection tour in the original high-dimensional space (using the \texttt{langevitour} package), displaying both the data and the model. The function takes the output from \texttt{comb\_all\_data\_model\_error()} (\texttt{point\_data}) and \texttt{edge\_data} which defines connections between neighboring bins.

@@

Several core computations within \texttt{quollr} are optimized using compiled C++ code via the \texttt{Rcpp} and \texttt{RcppArmadillo} packages. While the user interacts with high-level R functions, performance-critical steps such as nearest-neighbor searches (\texttt{compute\_highd\_dist()}), error metrics (\texttt{compute\_errors()}), $2\text{-}D$ distance calculations (\texttt{calc\_2d\_dist\_cpp()}), and generation of hexagon coordinates (\texttt{gen\_hex\_coord\_cpp()}) are handled internally in C++. This design provides significant speedups when analyzing large datasets while maintaining a user-friendly R interface. These C++ functions are not exported but are bundled within the package and fully accessible for inspection in the source code.

@@

Limb muscle cells of mice in \citet{tabula2018} are examined. There are \(1067\) single cells, with \(14997\) gene expressions. Following their pre-processing, different NLDR methods were performed using ten principal components. Figure \ref{fig:limb-rmse} (b) is the reproduction of the published plot. The question is whether this accurately represents the cluster structure in the data. Our method help to provide a better $2\text{-}D$ layout.

@@

The \texttt{quollr} package introduces a new framework for interpreting NLDR outputs by fitting a geometric wireframe model in $2\text{-}D$ and lifting it into high-dimensional space. This lifted model provides a direct way to assess how well a $2\text{-}D$ layout, produced by methods such as tSNE, UMAP, PHATE, TriMAP, or PaCMAP, preserves the structure of the original high-dimensional data. The approach offers both numerical and visual diagnostics to support the selection of NLDR methods and tuning hyper-parameters that produce the most accurate $2\text{-}D$ representations.

@@

To support efficient computation, particularly for large-scale datasets, several core functions in \texttt{quollr} are implemented in C++ using \texttt{Rcpp} and \texttt{RcppArmadillo}. These include functions for computing Euclidean distances in high-dimensional and $2\text{-}D$ space, identifying nearest centroids, calculating residual errors, and generating polygonal coordinates of hexagons. For instance, \texttt{compute\_highd\_dist()} accelerates nearest neighbor lookup in high-dimensional space, \texttt{compute\_errors()} calculates RMSE and total absolute error efficiently, and \texttt{calc\_2d\_dist\_cpp()} speeds up distance calculations in $2\text{-}D$. Additionally, \texttt{gen\_hex\_coord\_cpp()} constructs the coordinates for hexagonal bins based on their centroids with minimal overhead. These optimizations result in substantial performance gains compared to native R implementations, making the package responsive even when used in interactive contexts or on large datasets such as single-cell transcriptomic profiles.
